---
title: "final_project_analysis"
authors: "Wyatt Sinclair, Aniroop Naladala, Abhinav Pillai, Alyvia Fager"
output: html_document
date: "2026-04-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  warning = FALSE)

library(tidyverse)
library(janitor)
library(scales)
library(knitr)
library(corrplot)
```

```{r Data Loading}
cbb_raw <- read_csv("../data/cbb.csv")

glimpse(cbb_raw)
```

**Question 1:** Which team statistics are mostly strongly associated with wins?

```{r}
# Clean the data and create variables
cbb_clean <- cbb_raw %>%
  clean_names() %>%
  mutate(
    win_pct = w / g,
    efficiency_margin = adjoe - adjde
  ) %>%
  select(win_pct, adjoe, adjde, adj_t, barthag, efg_o, efg_d, efficiency_margin) %>%
  drop_na()

# Correlation matrix
numeric_vars <- cbb_clean %>%
  select(win_pct, adjoe, adjde, adj_t, barthag, efg_o, efg_d, efficiency_margin)

cor_matrix <- cor(numeric_vars)

cor_matrix
```

The correlation table provides numerical evidence of the relationships between team statistics and win percentage. Consistent with the heatmap, efficiency margin shows the strongest positive relationship with winning (0.76), indicating that teams that outscore their opponents by larger margins tend to be more successful. Offensive efficiency (0.68) also has a strong positive association with win percentage, while defensive efficiency (−0.62) has a strong negative relationship, meaning that teams that allow fewer points perform better.

Additionally, BARTHAG has a high correlation with both win percentage (0.75) and efficiency margin (0.99), reinforcing its role as a comprehensive measure of team strength. Shooting efficiency metrics, such as effective field goal percentage on offense (0.61) and defense (−0.58), also show meaningful relationships with winning. In contrast, adjusted tempo has almost no correlation with win percentage (−0.01), suggesting that the pace of play does not significantly influence team success.

```{r}
# correlation plot
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  col = colorRampPalette(c("red", "white", "blue"))(200),
  addCoef.col = "black",   # shows correlation numbers
  tl.col = "black",        # axis label color
  tl.srt = 45              # rotate labels
)
```

**Figure 1:** Correlation between team statistics and win percentage

The heatmap shows that efficiency margin has the strongest relationship with win percentage (0.76), indicating that teams that outscore opponents by larger margins win more games. Offensive efficiency (0.68) and defensive efficiency (−0.62) are both strongly related to winning, with offense showing a slightly stronger association. BARTHAG is also highly correlated with both efficiency margin and win percentage, reinforcing its role as a composite performance metric. In contrast, tempo exhibits near-zero correlation, suggesting that pace of play does not significantly influence team success. Overall these results suggest that team success is driven more by efficiency based performance than by style of play.

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
